---
title: "RNCBIGene: NCBI Gene annotation in parquet"
shorttitle: "NCBI Gene for all organisms"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{RNCBIGene: NCBI Gene annotation in parquet}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

Bioconductor's annotation system has functioned for well
over a decade as a pivotal resource for analysis and
tool development.  See the [AnnotationDbi](https://bioconductor.org/packages/AnnotationDbi)
documentation for details.  AnnotationDbi and packages based
on it use SQLite to manage
organism-specific vocabularies and to resolve queries about
mappings between identifier sets.  For example, learn the Gene Ontology
annotations for gene ORMDL3 in _Homo sapiens_:

```{r lkorg1, message=FALSE}
library(org.Hs.eg.db)
go_orm = AnnotationDbi::select(org.Hs.eg.db, key="ORMDL3", keytype="SYMBOL",
    columns=c("GO", "ENTREZID"))
head(go_orm)
```

RNCBIGene takes a different approach: instead of organism-specific SQLite
databases, all NCBI Gene annotation for all organisms lives in a set of
Apache Parquet files on an NSF Open Storage Network bucket.  Queries run
through a persistent duckdb connection that reads remote parquet directly,
pushing filters and column selections to the network layer before any data
lands in R.  The `open_ncbi_gene()` function returns a lazy `dplyr::tbl`;
the user composes `filter()`, `select()`, and `collect()` calls as they would
with any remote dplyr backend.

The duckdb `httpfs` extension required for remote parquet access is cached
permanently in `tools::R_user_dir("RNCBIGene", "data")` rather than a
temporary directory, so it is downloaded only once per machine.  The cache
location can be overridden by setting the `DUCKDB_EXTENSION_DIRECTORY`
environment variable before loading the package.

```{r lknc1, message=FALSE}
library(RNCBIGene)
library(dplyr)
lkor = open_ncbi_gene("gene2go", taxid=9606L) |>
  filter(GeneID==94103) |> collect() |> as.data.frame()
DT::datatable(lkor)
```

It is not the purpose of this package to provide pin-compatible
replacements for the AnnotationDbi packages.  This package explores
the opportunities for simplification and performance enhancement arising
from the adoption of Parquet and Arrow for annotation representation
and interrogation.

# Scope

There are eight parquet files representing NCBI Gene content.  Seven were
built from May 2025 NCBI snapshots; `gene2ensembl` was added to support
clean Ensembl identifier lookups.

```{r lksev}
tryCatch(
  available_ncbi_parquet(),
  error = function(e) message(conditionMessage(e))
)
```

All resources can be queried simultaneously through the persistent duckdb
connection.  Record counts are retrieved by pushing `COUNT(*)` to duckdb,
which reads row-group metadata from parquet without fetching the data itself.

```{r lksim}
tryCatch({
  resources <- available_ncbi_parquet()
  counts <- vapply(
    sub("\\.parquet$", "", resources),
    function(r) open_ncbi_gene(r) |> dplyr::tally() |> dplyr::collect() |> dplyr::pull(n),
    numeric(1))
  setNames(counts, resources)
}, error = function(e) message(conditionMessage(e)))
```

For illustration, a small subset covering four human genes on chromosome 17
is bundled with the package.

```{r lklit}
litp = dir(system.file("litparquet", package="RNCBIGene"), full.names=TRUE)
litpop = lapply(litp, arrow::open_dataset)
ns = basename(litp)
names(litpop) = ns
sapply(litpop, nrow)
```

Find the UniProt identifiers associated with ORMDL3:

```{r lkuni}
pro = litpop[["lit.gene2accession.parquet"]] |>
   filter(`#tax_id` == 9606, Symbol == "ORMDL3") |>
   select(protein_accession.version) |> collect() |> pull() |> setdiff("-")
litpop[["lit.gene_refseq_uniprotkb_collab.parquet"]] |>
   filter(NCBI_tax_id == 9606) |>
   filter(`#NCBI_protein_accession` %in% pro) |> collect()
```

# Lazy remote queries with `open_ncbi_gene()`

`open_ncbi_gene()` returns a lazy `dplyr::tbl` backed by a duckdb VIEW over
the remote parquet file.  No data is fetched until `collect()` is called.
Predicates passed to `filter()` and column lists in `select()` are translated
to SQL and pushed down to the parquet read, so only the requested rows and
columns travel the wire.

```{r lkopen}
tbl <- open_ncbi_gene("gene_info", taxid = 9606L)
class(tbl)
```

The object is a lazy SQL-backed tbl — inspecting it triggers a small preview
fetch but does not load the full dataset.

Compose a multi-step query before collecting:

```{r lkchain}
open_ncbi_gene("gene_info", taxid = 9606L) |>
  filter(Symbol %in% c("TP53", "ORMDL3", "BRCA1", "GSDMB")) |>
  select(Symbol, GeneID, map_location, type_of_gene) |>
  collect()
```

The same interface works for any of the eight resources.  Omitting `taxid`
returns a tbl spanning all organisms:

```{r lkortho, eval=FALSE}
open_ncbi_gene("gene_orthologs") |>
  filter(GeneID == 7157L) |>   # TP53 orthologs across all taxa
  collect()
```

# Discovering available fields with `ncbi_gene_fields()`

Before composing a query it is useful to know what columns a resource
provides.  `ncbi_gene_fields()` queries the duckdb schema for any resource and
returns a data frame of column names and their duckdb types — no data rows are
fetched.

```{r lkfields_info}
ncbi_gene_fields("gene_info")
```

The function accepts any name returned by `available_ncbi_parquet()`, with or
without the `.parquet` suffix:

```{r lkfields_go}
ncbi_gene_fields("gene2go")
```

Column names can also be used to validate a `select()` or `by` argument before
running a potentially expensive query:

```{r lkfields_check}
"map_location" %in% ncbi_gene_fields("gene_info")$column_name
"map_location" %in% ncbi_gene_fields("gene2go")$column_name
```

# Identifier mapping with `mapIdsNG()`

`mapIdsNG()` is a `mapIds`-style function that pushes all filtering to duckdb.
For Symbol or GeneID keytypes the filter is a SQL `IN (...)` predicate on
`gene_info`; for the Ensembl keytype a JOIN between `gene2ensembl` and
`gene_info` is performed entirely in duckdb.  Exactly one `collect()` call
retrieves only the matched rows and the requested columns.

## Symbol to GeneID

```{r lkmapids}
mapIdsNG(
  keys    = c("ORMDL3", "TP53", "GSDMB", "XyZZY"),
  keytype = "Symbol",
  column  = "GeneID"
)
```

The result is a named vector conformant to the input `keys`, with `NA` for
unrecognised identifiers.

## Symbol to chromosomal map location

```{r lkmaploc}
mapIdsNG(
  keys    = c("ORMDL3", "TP53", "GSDMB", "BRCA1"),
  keytype = "Symbol",
  column  = "map_location"
)
```

## Ensembl to Symbol

Ensembl lookups join `gene2ensembl.parquet` with `gene_info.parquet` on
`GeneID`; the join runs entirely in duckdb before a single `collect()`.

```{r lkensembl}
mapIdsNG(
  keys    = c("ENSG00000073605", "ENSG00000141510", "ENSG00000012048"),
  keytype = "Ensembl",
  column  = "Symbol"
)
```

# Querying non-human organisms

Because all NCBI Gene annotation for all organisms lives in the same parquet
files, switching species is simply a matter of changing `taxid`.  Here we look
up Ensembl gene and transcript identifiers for the mouse (_Mus musculus_,
taxid 10090) gene _Lilrb4a_.

A quick symbol-to-Ensembl-gene lookup:

```{r lkmouse1}
mapIdsNG(keys = "Lilrb4a", keytype = "Symbol", column = "Ensembl",
         taxid = 10090L)
```

For the full transcript-level picture we join `gene_info` (filtered to the
symbol) against `gene2ensembl` entirely within duckdb — both lazy tbls share
the same connection, so the join is pushed down before any data is collected:

```{r lkmouse2}
gene_info_mouse <- open_ncbi_gene("gene_info", taxid = 10090L) |>
  dplyr::filter(.data[["Symbol"]] == "Lilrb4a") |>
  dplyr::select(GeneID, Symbol)

open_ncbi_gene("gene2ensembl", taxid = 10090L) |>
  dplyr::inner_join(gene_info_mouse, by = "GeneID") |>
  dplyr::select(Symbol, Ensembl_gene_identifier,
                Ensembl_rna_identifier, Ensembl_protein_identifier) |>
  dplyr::collect()
```

The result reveals two distinct Ensembl gene models for _Lilrb4a_ — something
that a two-step approach (Symbol → one gene ID → transcripts) would have
missed.  Joining directly on `GeneID` returns all Ensembl records for the gene
in a single query.

# Joining local data to remote annotation with `join_ncbi_gene()`

`join_ncbi_gene()` uploads a local data frame to the duckdb session as a
temporary table and returns a lazy join against any remote NCBI Gene resource.
Both sides of the join run in duckdb; the caller decides what to `select()` and
when to `collect()`.

```{r lkjoin}
local_df <- data.frame(Symbol = c("ORMDL3", "TP53", "BRCA1", "GSDMB", "XyZZY"))

join_ncbi_gene(local_df, by = "Symbol", taxid = 9606L) |>
  select(Symbol, GeneID, map_location, type_of_gene) |>
  collect()
```

Unmatched keys (here `XyZZY`) appear with `NA` in the annotation columns
because the default join type is `"left"`.  Switch to `type = "inner"` to drop
them.

## Enriching a GeneID table with GO terms and RefSeq identifiers

When multiple annotation resources are needed, each requires a **separate**
join back to the original data frame.  GO terms and RefSeq identifiers are
both one-to-many relationships with GeneID: three genes yield hundreds of GO
annotations and hundreds of transcript/protein pairs.  Joining both resources
in a single query would produce a Cartesian product; the correct pattern is two
independent joins.

Start with a local data frame of Entrez IDs:

```{r lkjoin2}
local_ids <- data.frame(GeneID = c(94103L, 7157L, 672L))
```

**GO terms** — one row per gene-GO pair:

```{r lkjoin_go}
go_tbl <- join_ncbi_gene(local_ids, by = "GeneID",
                          resource = "gene2go", taxid = 9606L) |>
  select(GeneID, GO_ID, GO_term, Evidence, Category) |>
  collect()
nrow(go_tbl)
head(go_tbl)
```

**RefSeq identifiers** — one row per gene-transcript pair (excluding entries
with no RNA accession):

```{r lkjoin_refseq}
refseq_tbl <- join_ncbi_gene(local_ids, by = "GeneID",
                              resource = "gene2refseq", taxid = 9606L) |>
  select(GeneID,
         RNA   = RNA_nucleotide_accession.version,
         protein = protein_accession.version) |>
  filter(RNA != "-") |>
  collect()
nrow(refseq_tbl)
head(refseq_tbl)
```

Each result shares the `GeneID` key with the original `local_ids` data frame
and can be used independently or merged with `dplyr::left_join()` after
collecting, depending on the downstream analysis.

# Session information

```{r lksess}
sessionInfo()
```
